---
title: "DataViz"
author: "Mia Forsline"
date: "8/18/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#general use
library(tidyverse)
library(here)
library(janitor)

#data viz packages 
library(patchwork)
library(ggrepel)
library(gghighlight)
library(paletteer)
library(ggExtra)
library(ggbeeswarm)

#contains data
library(gapminder)

library(sf)
```

## Get the lizards data
```{r}
#use getwd() or here() to identify working directory 
lizards <- read_csv(here("data_tidy", "lizards.csv"))
```

## Histogram of lizard weight
```{r}
ggplot(data = lizards, aes(x = weight))+
  geom_histogram(fill = "lavender", 
                 color = "mediumpurple4",
                 size = 0.75,
                 linetype = "dashed")+
  theme_classic()+
  labs(x = "Lizard Weight (g)")
```
## Scatterplot 
```{r}
#shape = 0 - 25 
ggplot(data = lizards, aes(x = total_length, y = weight))+
  geom_point(alpha = 0.5, 
             shape = 100, 
             color = "lightcoral",
             size = 3)+
  theme_classic()+
  labs(x = "Lizard Length", 
       y = "Lizard Weight (g)")
```
## Bubble plot 
```{r}
ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point(aes(color = common_name, 
                 size = total_length), 
             alpha = 0.25,
             shape = 18) +
  theme(legend.position = "none")
```

```{r}
#facet_grid = a true grid that will create a combination of the intersections of the variables 
#facet_wrap = puts all the graphs in a row then wraps to the next line 

ggplot(data = lizards, aes(x = total_length, y = weight))+ 
  geom_point()+
  facet_wrap(.~common_name, ncol = 2) #can choose how many columns before wrapping to the next line

ggplot(data = lizards, aes(x = total_length, y = weight))+ 
  geom_point(aes(color = sex, shape = tail), alpha = 0.5)+
  facet_grid(sex ~ tail) #rows ~ columns
#populates all intersections of the grid 
```

## ggplot default order for characters is alphabetical - how can we choose other orders ? 
### Find total lizard counts by common name 
```{r}
lizard_counts <- lizards %>% 
  group_by(common_name) %>% 
  summarize(count = n())

lizard_counts

#another way to find the count 
lizard_counts <- lizards %>% dplyr::count(common_name)

lizard_counts_2 <- lizards %>% count(common_name, tail) #can group counts by multiple variables 
```

```{r}
ggplot(data = lizard_counts, aes(y = fct_reorder(common_name, count), x = count))+ #long y-axis common names are readable with sideways columns 
  geom_col()+
  theme_classic()+
  labs(y = "Lizard Common Names", 
       x = "Count")

#can use coord_flip() to switch x and y axes OR just specify x and y axes 

#it's not logical to have the lizard names in alphabetical order (which is the default)
#factors = character/number with meaningful order assigned to discrete groups 
#so let's turn the names into factors and have their levels assigned based on their count values so we can order them from low to high 
#forcats::fct_reorder(object to convert, external variable to base the level based on) to convert to a factor and reorder the levels based on a value in another variable 
#forcats is part of the tidyverse 
```

Try converting common_name to an ordered factor outside of ggplot 
```{r}
lizard_counts_new <- lizard_counts %>% 
  mutate(common_name = fct_reorder(common_name, count))
#can check new factor's levels using levels()
```

## Axis scale breaks, limits, and labels 
```{r}
ggplot(data = lizards, aes(x = total_length, y = weight))+
  geom_point(alpha = 0.5, color = "dodgerblue4")+
  theme_classic()+
  scale_x_continuous(breaks = c(0, 10, 50, 5000),
                     lim = c(0, 500),
                     expand = c(0, 0)) + #prevent expansion beyond limits 
  scale_y_continuous(lim = c(min(lizards$weight), max(lizards$weight)))
  
#what type of variables are on the axes ? continues, discrete, or date ?
#breaks = tick mark breaks. you can create the vector manually, using seq(), etc. 
```
Transform the date column to class Date then find counts by date 
```{r}
lizards_counts <- lizards %>% 
  mutate(date = lubridate::mdy(date)) %>% #either attach library(lubridate) or specify the package
  group_by(date) %>% 
  summarize(count = n())

lizards_counts
```

```{r}
ggplot(data = lizards_counts, aes(x = date, y = count))+
  geom_line()+
  theme_classic()+
  labs(x = "Time", 
       y = "Lizard Frequency")+
  scale_x_date(breaks = scales::breaks_width("3 years"), #or scale_x_date(date_breaks = "3 years")
               labels = scales::label_date("%y"))  #or date_labels = "%y"
```
## Scale colors 
```{r}
ggplot(data = lizards, aes(x = total_length, y = weight)) + 
  geom_point(aes(color = weight)) +
  #scale_color_gradient(low = "red", high = "navy") choose low and high continuous gradient colors
  #scale_color_gradientn(colors = c("orange", "purple", "red", "black")) #choose more than 2 gradient colors 
  scale_color_stepsn(colors = c("green", "blue", "purple", "pink"),
                     breaks = seq(from = 0, to = 100, by = 20)) #color steps rather than a gradient 
```
### Update a color scheme using a palette in paletteer
```{r}
lizards_fct <- lizards %>% 
  mutate(common_name = fct_reorder(.f = common_name, .x = total_length, .fun = median)) #convert common_name to a factor ordered by total length median 

ggplot(data = lizards_fct, aes(y = common_name, x = total_length))+
  geom_boxplot(aes(fill = common_name), show.legend = FALSE)+
  #geom_jitter(alpha = 0.25, aes(color = common_name))+
  theme_classic()+
  scale_fill_paletteer_d(palette = "palettetown::baltoy") #select paleteer package::palette
  
#View(palettes_d_names)
```

```{r}
ggplot(data = lizards, aes(x = total_length, y = weight))+
  geom_point()+
  theme(
    #panel.background = element_rect(fill = "yellow", 
    #                                color = "purple",
    #                                size = 10),
    #panel.grid.major.x = element_line(color = "red"), #applies to x gridlines only 
    #panel.grid = element_line(color = "blue"), #applies to x and y gridlines 
    #panel.grid.major.y = element_line(color = "orange"),
    panel.grid = element_blank(), #remove gridlines 
    
    plot.background = element_rect(fill = "cyan"), #fill background panel
    
    axis.text.x = element_text(color = "orange"),
    
    axis.title = element_text(color = "yellow", size = 15)
  )

#most specific level > general commands
```

## ggrepel - for labeling plots 
```{r}
ww_lizards <- lizards %>% 
  filter(common_name == "western whiptail" & site == "sand")

ggplot(data = ww_lizards, aes(x = total_length, y = weight))+ 
  geom_point()+
  geom_text_repel(aes(label = toe_num), #add label text to each point but not on top of each point 
                  size = 3,
                  max.overlaps = 20) 
```

### use gapminder data 
```{r}
gap <- gapminder %>% 
  filter(continent == "Europe" & year == "2007")

ggplot(data = gap, aes(x = gdpPercap, y = lifeExp))+
  geom_point()+
  geom_text_repel(aes(label = country), 
                  size = 3, 
                  max.overlap = 30)+
  theme_classic()
```

## gghighlight - highlight data that meet conditions you set 
```{r}
p <- ggplot(lizards, aes(x = total_length, y = weight))+
  geom_point()

p + gghighlight(toe_num == 250, label_key = toe_num)
```

```{r}
q <- ggplot(lizards, aes(x = total_length, y = weight))+
  geom_line(aes(color = common_name))

q + gghighlight(max(weight) > 30)
```

## patchwork for compound figures 
```{r}
p | q #puts both graphs side by side using patchwork 

(p | q) / q #use PEMDAS to place p next to q all over q 

p / q / p & theme_classic() #use & to apply themes to all figures 

#can use calPlot package in addition to patchwork 
```

## A few new graph types
### Marginal plots using ggExtra()
```{r}
whiptails <- lizards %>% 
  filter(common_name == "western whiptail") %>% 
  drop_na(total_length, weight) #delete complete row if NAs are present in total_length or weight

ggplot(data = whiptails, aes(x = total_length, y = weight))+
  geom_point()+
  geom_rug() #rug plots = puts tickmarks at any value in order to see distribution of values. not always useful because it doesn't show density at each increment 

```
### Marginal plot w/boxplots
```{r}
my_plot <- ggplot(data = whiptails, aes(x = total_length, y = weight))+
  geom_point(aes(color = sex), size = 2)+
  scale_color_manual(values = c("cyan4", "black", "goldenrod"),
                     name = "Sex:", 
                     labels = c("female", "juvenile", "male"))+
  
  theme_classic()+
  theme(legend.position = "bottom")

ggMarginal(my_plot, type = "boxplot", groupColour = TRUE)
  
```

## Beeswarm plot 
```{r}
ggplot(data = whiptails, aes(x = sex, y = weight))+
  geom_beeswarm()+
  #geom_violin(alpha = 0.5)+
  geom_boxplot(alpha = 0.5)
```

## A heat map w/geom_tile()
```{r}
lizard_counts <- lizards %>% 
  mutate(date = lubridate::mdy(date)) %>% #convert to class date
  mutate(year = lubridate::year(date)) %>% #pull the year from the date and create a new column 
  group_by(year, common_name) %>% 
  summarize(count = n())
#can also use count(year, common_name) instead of group_by() + summarize(n())
lizard_counts
```
#3D plot - showing year, common name, and count 
```{r}
ggplot(data = lizard_counts, aes(x = year, y = common_name))+
  geom_tile(aes(fill = count)) +
  geom_text(aes(label = count), color = "white", size = 3) + #we want the label to be in the middle of each tile, so no need for teom_text_repel
  #scale_fill_gradientn(colors = c("goldenrod", "orange", "darkred"))+ #color gradient with as many colors as we want 
  scale_fill_viridis_c() #continuous scale color-blind-friendly colors 
```

### Make a map using `sf` (simple features) to work w/spatial data 
```{r}
jornada_veg <- read_sf(here("data_raw", "spatial_vegetation", "doc.kml")) %>% 
  select(Name) %>%  #sticky geometry column stays even when we select Name col
  clean_names()

ggplot(data = jornada_veg)+
  geom_sf(aes(fill = name),  #plot simple features 
          color = NA) + #can change borders around polygons 
  scale_fill_paletteer_d(palette = "palettetown::vaporeon") + #discrete 
  labs(x = "Longitude", 
       y = "Latitude", 
       fill = "Dominant vegetation")+
  theme_void()
```

