---
title: "DataViz"
author: "Mia Forsline"
date: "8/18/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#general use
library(tidyverse)
library(here)
library(janitor)

#data viz packages 
library(patchwork)
library(ggrepel)
library(gghighlight)
library(paletteer)
library(ggExtra)
library(ggbeeswarm)

#contains data
library(gapminder)

library(sf)
```

## Get the lizards data
```{r}
#use getwd() or here() to identify working directory 
lizards <- read_csv(here("data_tidy", "lizards.csv"))
```

## Histogram of lizard weight
```{r}
ggplot(data = lizards, aes(x = weight))+
  geom_histogram(fill = "lavender", 
                 color = "mediumpurple4",
                 size = 0.75,
                 linetype = "dashed")+
  theme_classic()+
  labs(x = "Lizard Weight (g)")
```
## Scatterplot 
```{r}
#shape = 0 - 25 
ggplot(data = lizards, aes(x = total_length, y = weight))+
  geom_point(alpha = 0.5, 
             shape = 100, 
             color = "lightcoral",
             size = 3)+
  theme_classic()+
  labs(x = "Lizard Length", 
       y = "Lizard Weight (g)")
```
## Bubble plot 
```{r}
ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point(aes(color = common_name, 
                 size = total_length), 
             alpha = 0.25,
             shape = 18) +
  theme(legend.position = "none")
```

```{r}
#facet_grid = a true grid that will create a combination of the intersections of the variables 
#facet_wrap = puts all the graphs in a row then wraps to the next line 

ggplot(data = lizards, aes(x = total_length, y = weight))+ 
  geom_point()+
  facet_wrap(.~common_name, ncol = 2) #can choose how many columns before wrapping to the next line

ggplot(data = lizards, aes(x = total_length, y = weight))+ 
  geom_point(aes(color = sex, shape = tail), alpha = 0.5)+
  facet_grid(sex ~ tail) #rows ~ columns
#populates all intersections of the grid 
```

## ggplot default order for characters is alphabetical - how can we choose other orders ? 
### Find total lizard counts by common name 
```{r}
lizard_counts <- lizards %>% 
  group_by(common_name) %>% 
  summarize(count = n())

lizard_counts

#another way to find the count 
lizard_counts <- lizards %>% dplyr::count(common_name)

lizard_counts_2 <- lizards %>% count(common_name, tail) #can group counts by multiple variables 
```

```{r}
ggplot(data = lizard_counts, aes(y = fct_reorder(common_name, count), x = count))+ #long y-axis common names are readable with sideways columns 
  geom_col()+
  theme_classic()+
  labs(y = "Lizard Common Names", 
       x = "Count")

#can use coord_flip() to switch x and y axes OR just specify x and y axes 

#it's not logical to have the lizard names in alphabetical order (which is the default)
#factors = character/number with meaningful order assigned to discrete groups 
#so let's turn the names into factors and have their levels assigned based on their count values so we can order them from low to high 
#forcats::fct_reorder(object to convert, external variable to base the level based on) to convert to a factor and reorder the levels based on a value in another variable 
#forcats is part of the tidyverse 
```

Try converting common_name to an ordered factor outside of ggplot 
```{r}
lizard_counts_new <- lizard_counts %>% 
  mutate(common_name = fct_reorder(common_name, count))
#can check new factor's levels using levels()
```

## Axis scale breaks, limits, and labels 
```{r}
ggplot(data = lizards, aes(x = total_length, y = weight))+
  geom_point(alpha = 0.5, color = "dodgerblue4")+
  theme_classic()+
  scale_x_continuous(breaks = c(0, 10, 50, 5000),
                     lim = c(0, 500),
                     expand = c(0, 0)) + #prevent expansion beyond limits 
  scale_y_continuous(lim = c(min(lizards$weight), max(lizards$weight)))
  
#what type of variables are on the axes ? continues, discrete, or date ?
#breaks = tick mark breaks. you can create the vector manually, using seq(), etc. 
```
Transform the date column to class Date then find counts by date 
```{r}
lizards_counts <- lizards %>% 
  mutate(date = lubridate::mdy(date)) %>% #either attach library(lubridate) or specify the package
  group_by(date) %>% 
  summarize(count = n())

lizards_counts
```

```{r}
ggplot(data = lizards_counts, aes(x = date, y = count))+
  geom_line()+
  theme_classic()+
  labs(x = "Time", 
       y = "Lizard Frequency")+
  scale_x_date(breaks = scales::breaks_width("3 years"), #or scale_x_date(date_breaks = "3 years")
               labels = scales::label_date("%y"))  #or date_labels = "%y"
```
## Scale colors 
```{r}
ggplot(data = lizards, aes(x = total_length, y = weight)) + 
  geom_point(aes(color = weight)) +
  #scale_color_gradient(low = "red", high = "navy") choose low and high continuous gradient colors
  #scale_color_gradientn(colors = c("orange", "purple", "red", "black")) #choose more than 2 gradient colors 
  scale_color_stepsn(colors = c("green", "blue", "purple", "pink"),
                     breaks = seq(from = 0, to = 100, by = 20)) #color steps rather than a gradient 
```
### Update a color scheme using a palette in paletteer
```{r}
lizards_fct <- lizards %>% 
  mutate(common_name = fct_reorder(.f = common_name, .x = total_length, .fun = median)) #convert common_name to a factor ordered by total length median 

ggplot(data = lizards_fct, aes(y = common_name, x = total_length))+
  geom_boxplot(aes(fill = common_name), show.legend = FALSE)+
  #geom_jitter(alpha = 0.25, aes(color = common_name))+
  theme_classic()+
  scale_fill_paletteer_d(palette = "palettetown::baltoy") #select paleteer package::palette
  
#View(palettes_d_names)
```

```{r}
ggplot(data = lizards, aes(x = total_length, y = weight))+
  geom_point()+
  theme(
    #panel.background = element_rect(fill = "yellow", 
    #                                color = "purple",
    #                                size = 10),
    #panel.grid.major.x = element_line(color = "red"), #applies to x gridlines only 
    #panel.grid = element_line(color = "blue"), #applies to x and y gridlines 
    #panel.grid.major.y = element_line(color = "orange"),
    panel.grid = element_blank(), #remove gridlines 
    
    plot.background = element_rect(fill = "cyan"), #fill background panel
    
    axis.text.x = element_text(color = "orange"),
    
    axis.title = element_text(color = "yellow", size = 15)
  )

#most specific level > general commands
```

